---
title: "Squamous Tumor DEG Analysis"
author: "Matthew Zatzman"
date: "`r Sys.Date()`"
format: 
  html:
    code-fold: true
    code-tools: true
    code-link: true
    toc: true
---

# Introduction

Reprocessing the squamous dataset

```{r setup}
#| output: false
library(tidyverse)
# library(zellkonverter)
library(ggthemes)
library(dittoSeq)
library(patchwork)
library(Matrix)
library(glue)
library(ggplotify)
library(ggh4x)
library(patchwork)
library(Seurat)
library(here)

options(Seurat.object.assay.version = "v5")
```

Load data

```{r}
srt <- readRDS("squamous_cohort_celltyped.rds")
```


## Macrophages

Pull out and subphenotype macrophages

```{r}
FeaturePlot(srt[,srt$cell_type_coarse == "Macrophage"], reduction = "umap_harmony", features = c("LYZ", "AIF1"))
```


```{r}
cell_col = "cell_type_coarse"
cell_type = "Macrophage"

srt$macrophage <- srt@meta.data %>%
  mutate(macrophage = case_when(
    cell_type_coarse == "Macrophage" & 
      predicted.ann_level_1 == "Immune" & 
      predicted.ann_level_2 == "Myeloid" &
      predicted.ann_level_3 %in% c("Dendritic cells", "Macrophages", "Monocytes") ~ TRUE,
    .default = F
  )) %>%
  pull(macrophage)

resolution = 0.5
k.param = 5
n.neighbors = 20
dims = 1:30

# subcluster_cells <- function(srt, cell_col, cell_type, k.param = 10, n.neighbors = 20, dims = 1:20, resolution = 1) {
  
  # srt_cls <- srt[,srt[[cell_col]] == cell_type] %>%
  srt_cls <- srt[,srt$macrophage] %>%
    FindVariableFeatures() %>%
    ScaleData(vars.to.regress = "percent.mt") %>%
    RunPCA() %>%
    IntegrateLayers(method = HarmonyIntegration, orig.reduction = "pca", new.reduction = "harmony", group.by = "sample_id", max.iter.harmony = 20) %>%
  FindNeighbors(reduction = "harmony", dims = dims, graph.name = c("harmony.nn", "harmony.snn"), k.param = k.param) %>%
  RunUMAP(dims = dims, reduction = "harmony", reduction.name = "umap_harmony", n.neighbors = n.neighbors) %>% 
  FindClusters(resolution = resolution, graph.name = "harmony.snn", cluster.name = "clusters_harmony")
  
  srt_cls$clusters_harmony_leiden <- factor(scatools::leiden_wrapper(srt_cls@graphs$harmony.snn, resolution = 0.5))

  
  # return(srt_cls)
# }
```


## Macrophage deep dive

```{r}
# macro <- subcluster_cells(srt, cell_col = "cell_type_coarse", cell_type = "Macrophage", k.param = 10, n.neighbors = 20, dims = 1:30, resolution = 0.4)
```

```{r}
DimPlot(srt[,srt$macrophage], reduction = "umap_harmony")
```


```{r}
macro <- srt_cls
p1 <- DimPlot(macro, reduction = "umap_harmony", group.by = "clusters_harmony_leiden", cols = dittoColors(), label = T, label.box = F)
p1
```

```{r}
Idents(macro) <- "clusters_harmony_leiden"
markers_macro <- FindAllMarkers(macro, only.pos = T, max.cells.per.ident = 500)
```


```{r, fig.width = 5, fig.height = 4}
Idents(macro) <- "clusters_harmony_leiden"

macro_markers <- c("LYZ", "AIF1", "FCER1G", "TYROBP", "MS4A6A")

feature_dotplot <- DotPlot(macro, features = macro_markers, assay = "RNA", scale = F, cluster.idents = F, group.by = "clusters_harmony_leiden", col.min = 0) + guides(x = guide_axis(angle = 90))
feature_dotplot + theme_classic() + theme(panel.spacing = unit(1, "mm"), strip.background.x = element_part_rect(side = "b"), panel.border = element_rect(fill = NA), strip.text.x = element_text(angle = 90, hjust = 0))

```


Plot top cluster markers

```{r, fig.width = 12, fig.height=5}
markers_plot <- markers_macro %>%
  mutate(fc_dir = sign(avg_log2FC)) %>%
  group_by(cluster) %>%
  slice_head(n = 6) %>%
  ungroup() %>%
  distinct(gene, .keep_all = T)
to_plot <- split(markers_plot$gene, markers_plot$cluster)


feature_dotplot <- DotPlot(macro, features = to_plot, assay = "RNA", scale = T, cluster.idents = F, group.by = "clusters_harmony_leiden", col.min = 0) + guides(x = guide_axis(angle = 90))
feature_dotplot + theme_classic() + theme(panel.spacing = unit(1, "mm"), strip.background.x = element_part_rect(side = "b"), panel.border = element_rect(fill = NA), strip.text.x = element_text(angle = 90, hjust = 0))
```

```{r, fig.width = 14, fig.height = 8}
p1 <- DimPlot(macro, reduction = "umap_harmony", group.by = "clusters_harmony_leiden", cols = dittoColors(), label = T, label.box = T)

Idents(macro) <- "clusters_harmony_leiden"

macro_markers <- c("LYZ", "AIF1", "FCER1G", "TYROBP", "MS4A6A")

feature_dotplot <- DotPlot(macro, features = macro_markers, assay = "RNA", scale = F, cluster.idents = F, group.by = "clusters_harmony_leiden", col.min = 0) + guides(x = guide_axis(angle = 90))
p2 <- feature_dotplot + theme_classic() + theme(panel.spacing = unit(1, "mm"), strip.background.x = element_part_rect(side = "b"), panel.border = element_rect(fill = NA), strip.text.x = element_text(angle = 90, hjust = 0))

markers_plot <- markers_macro %>%
  mutate(fc_dir = sign(avg_log2FC)) %>%
  group_by(cluster) %>%
  slice_head(n = 6) %>%
  ungroup() %>%
  distinct(gene, .keep_all = T)
to_plot <- split(markers_plot$gene, markers_plot$cluster)


feature_dotplot <- DotPlot(macro, features = to_plot, assay = "RNA", scale = T, cluster.idents = F, group.by = "clusters_harmony_leiden", col.min = 0) + guides(x = guide_axis(angle = 90))
p3 <- feature_dotplot + theme_classic() + theme(panel.spacing = unit(1, "mm"), strip.background.x = element_part_rect(side = "b"), panel.border = element_rect(fill = NA), strip.text.x = element_text(angle = 90, hjust = 0))


pcomb <- ((p1 | p2) / p3) + plot_annotation(title = "Macrophage compartment subclustering", theme = theme(plot.title = element_text(hjust = 0.5, size = 16, face = "bold")))
pcomb
```


```{r}
# macro[["percent.ribo"]] <- PercentageFeatureSet(macro, pattern = "^RP[SL]")
```


Dump out marker list (only sig)

```{r}
markers_macro %>%
  # filter(p_val_adj < 0.05) %>%
  write.table("macro_cluster_markers.txt", quote = F, sep = "\t", row.names = F, col.names = T)
```



```{r}
markers_new <-  c("VCAN", "SLC11A1", "S100A8", "S100A9", "CSF3R", "MARCO", "MAFB", "SPP1", "CD44", "ALOX5AP", "SELL", "PTGS2", "CCL20", "IL1RN", "CD14", "CD163", "TREM2", "PPARG")
```

```{r, fig.width = 8, fig.height=3}
feature_dotplot <- DotPlot(macro, features = markers_new, assay = "RNA", scale = F, cluster.idents = F, group.by = "clusters_harmony_leiden", col.min = 0) + guides(x = guide_axis(angle = 90))
feature_dotplot + theme_classic() + theme(panel.spacing = unit(1, "mm"), strip.background.x = element_part_rect(side = "b"), panel.border = element_rect(fill = NA), strip.text.x = element_text(angle = 90, hjust = 0))
```

Assign cell identities

```{r}
coarse_cluster_map_macro <- c(
  "1" = "Neutrophil? (exclude)",
  "2" = "MoMac 1",
  "3" = "Macrophage (classic)",
  "4" = "DC2",
  "5" = "Macrophage (complement)",
  "6" = "MoMac 2",
  "7" = "Macrophage (cycling)",
  "8" = "DCreg",
  "9" = "DC1",
  "10" = "Macrophage (MThi)",
  "11" = "KRThi (exclude)"
)


macro@meta.data$macrophage_subtype <- coarse_cluster_map_macro[match(macro@meta.data$clusters_harmony_leiden, names(coarse_cluster_map_macro))]

# srt$cell_type_coarse <- factor(srt$cell_type_coarse, levels = sort(unique(srt$cell_type_coarse)))
```


```{r, fig.width = 8, fig.height=3}
feature_dotplot <- DotPlot(macro, features = markers_new, assay = "RNA", scale = F, cluster.idents = F, group.by = "macrophage_subtype", col.min = 0) + guides(x = guide_axis(angle = 90))
feature_dotplot + theme_classic() + theme(panel.spacing = unit(1, "mm"), strip.background.x = element_part_rect(side = "b"), panel.border = element_rect(fill = NA), strip.text.x = element_text(angle = 90, hjust = 0))
```


```{r, fig.width = 14, fig.height = 8}
p1 <- DimPlot(macro, reduction = "umap_harmony", group.by = "macrophage_subtype", cols = dittoColors(), label = F, label.box = T)

Idents(macro) <- "clusters_harmony_leiden"

macro_markers <- c("LYZ", "AIF1", "FCER1G", "TYROBP", "MS4A6A")

feature_dotplot <- DotPlot(macro, features = macro_markers, assay = "RNA", scale = F, cluster.idents = F, group.by = "macrophage_subtype", col.min = 0) + guides(x = guide_axis(angle = 90))
p2 <- feature_dotplot + theme_classic() + theme(panel.spacing = unit(1, "mm"), strip.background.x = element_part_rect(side = "b"), panel.border = element_rect(fill = NA), strip.text.x = element_text(angle = 90, hjust = 0))

markers_plot <- markers_macro %>%
  mutate(fc_dir = sign(avg_log2FC)) %>%
  group_by(cluster) %>%
  slice_head(n = 6) %>%
  ungroup() %>%
  distinct(gene, .keep_all = T)
to_plot <- split(markers_plot$gene, markers_plot$cluster)


feature_dotplot <- DotPlot(macro, features = to_plot, assay = "RNA", scale = T, cluster.idents = F, group.by = "macrophage_subtype", col.min = 0) + guides(x = guide_axis(angle = 90))
p3 <- feature_dotplot + theme_classic() + theme(panel.spacing = unit(1, "mm"), strip.background.x = element_part_rect(side = "b"), panel.border = element_rect(fill = NA), strip.text.x = element_text(angle = 90, hjust = 0))


pcomb <- ((p1 | p2) / p3) + plot_annotation(title = "Macrophage compartment subclustering", theme = theme(plot.title = element_text(hjust = 0.5, size = 16, face = "bold")))
pcomb
```


```{r, fig.width = 11, fig.height=6}
#DC2 markers
FeaturePlot(macro, features = c("CD1C", "FCER1A", "PKIB", "CD68"), reduction = "umap_harmony", max.cutoff = "q99", order = T, label = T)
VlnPlot(macro, features = c("CD1C", "FCER1A", "PKIB"), group.by = "macrophage_subtype")
```

Coarse macrophage clusters

```{r}
# Remove the excluded stuff
macro <- macro[,!grepl("exclude", x = macro$macrophage_subtype)]

macro$macrophage_subtype_coarse <- macro@meta.data %>%
  mutate(macrophage_subtype_coarse = case_when(
    grepl("cycling", macrophage_subtype) ~ "Macrophage (cycling)",
    grepl("Macrophage", macrophage_subtype) ~ "Macrophage",
    .default = macrophage_subtype
  )) %>%
  pull(macrophage_subtype_coarse) %>%
  factor()

```

Coarse markers

```{r}
Idents(macro) <- "macrophage_subtype_coarse"
markers_macro_coarse <- FindAllMarkers(macro, only.pos = T, max.cells.per.ident = 500)
```

```{r}
markers_macro_coarse %>%
  # filter(p_val_adj < 0.05) %>%
  write.table("macro_cluster_markers_coarse.txt", quote = F, sep = "\t", row.names = F, col.names = T)
```


```{r, fig.width = 11, fig.height = 8}
p1 <- DimPlot(macro, reduction = "umap_harmony", group.by = "macrophage_subtype_coarse", cols = dittoColors(), label = F, label.box = T)

Idents(macro) <- "macrophage_subtype_coarse"

macro_markers <- c("LYZ", "AIF1", "FCER1G", "TYROBP", "MS4A6A")

feature_dotplot <- DotPlot(macro, features = macro_markers, assay = "RNA", scale = F, cluster.idents = F, group.by = "macrophage_subtype_coarse", col.min = 0) + guides(x = guide_axis(angle = 90))
p2 <- feature_dotplot + theme_classic() + theme(panel.spacing = unit(1, "mm"), strip.background.x = element_part_rect(side = "b"), panel.border = element_rect(fill = NA), strip.text.x = element_text(angle = 90, hjust = 0))

markers_plot <- markers_macro_coarse %>%
  mutate(fc_dir = sign(avg_log2FC)) %>%
  group_by(cluster) %>%
  slice_head(n = 6) %>%
  ungroup() %>%
  distinct(gene, .keep_all = T)
to_plot <- split(markers_plot$gene, markers_plot$cluster)

# to_plot$DCreg <- append(to_plot$DCreg, "CCR7")

feature_dotplot <- DotPlot(macro, features = to_plot, assay = "RNA", scale = T, cluster.idents = F, group.by = "macrophage_subtype_coarse", col.min = 0) + guides(x = guide_axis(angle = 90))
p3 <- feature_dotplot + theme_classic() + theme(panel.spacing = unit(1, "mm"), strip.background.x = element_part_rect(side = "b"), panel.border = element_rect(fill = NA), strip.text.x = element_text(angle = 90, hjust = 0))


pcomb <- ((p1 | p2) / p3) + plot_annotation(title = "Macrophage compartment clustering", theme = theme(plot.title = element_text(hjust = 0.5, size = 16, face = "bold")))
pcomb
```

Save out objects and calls

```{r}
saveRDS(macro, file = "squamous_cohort_celltyped_macro.rds")

# Convert to h5ad and save
sce <- as.SingleCellExperiment(macro)
sce$ident <- NULL

adata <- zellkonverter::SCE2AnnData(sce, X_name = "logcounts")
anndata::write_h5ad(adata, filename = "squamous_cohort_celltyped_macro.h5ad")
```



### TODO

1. Score metaprograms in the macrophages
2. CITE analysis (CD4 CD8 T cells)
    * Check Mac2 (neutrophil?) cluster
3. Revisit cell interaction analysis

